---
title: "Econ 104 Group Project 1"
author: "Robert Priolo"
date: "4/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list=ls())
t.critical <- function(alpha, df, tail = "two") {
  if(tail == "two"){
    result <- qt(1-alpha/2  , df)
  } else if(tail == "right"){
    result <- qt(1-alpha  , df)
  } else if(tail=="left"){
    result <- qt(alpha  , df)
  }
  return(result)
}
library(POE5Rdata)
data("vegas5")
attach(vegas5)

library(broom)
```


#Question 1

Provide a descriptive analysis of your variables. This should include histograms and fitted
distributions, correlation plot, boxplots, scatterplots, and statistical summaries (e.g., the
five-number summary). All figures must include comments.
```{r}
# Obs: 10,000 mortgage loan observations from Las Vegas, Nevada, single family homes, 2010
# 
# The authors would like to thank Meagan McCollum, Department of Finance, Louisiana State University for
# providing the data.
# 
# default         = 1 if payment late by 90+ days
# arm             = 1 if adjustable rate mortgage, 0 if fixed
# refinance       = 1 if loan is for a refinance of any type, 0 if for purchase
# lien2           = 1 if 2nd lien mortage, 0 if 1st lien
# term30          = 1 if 30 year mortgage, 0 if 15 year mortgage
# underwater      = 1 if borrower estimated to owe more than property worth at time of obs., 0 oth
# ltv             loan to value ratio of property at origination, percent
# rate            current interest rate on loan, percent
# amount          loan amount in $10,000 units
# fico            borrower's credit score at origination
# 
#     Variable |       Obs        Mean    Std. Dev.       Min        Max
# -------------+--------------------------------------------------------
#      default |     10000       .3699    .4828015          0          1
#          arm |     10000       .3783    .4849872          0          1
#    refinance |     10000       .5451    .4979867          0          1
#        lien2 |     10000       .1035     .304626          0          1
#       term30 |     10000       .8535    .3536245          0          1
# -------------+--------------------------------------------------------
#   underwater |     10000       .8228     .381857          0          1
#          ltv |     10000    69.25543    21.03323          5        109
#         rate |     10000    5.979238    2.063461         .5     17.375
#       amount |     10000    24.56128    20.69574     1.0948        750
#         fico |     10000    684.7314    66.11495        442        823


hist(default)
hist(arm)
hist(refinance)
hist(lien2)
hist(term30)
hist(underwater)
hist(ltv)
hist(rate)
hist(amount)
hist(fico)


```


#Question 2

Estimate a multiple linear regression model that includes all the main effects only (i.e., no
interactions nor higher order terms). We will use this model as a baseline. Comment on
the statistical and economic significance of your estimates. Also, make sure to provide an
interpretation of your estimates.

```{r}

model<- lm(default~arm+refinance+lien2+term30+underwater+ltv+rate+amount+fico)
summary(model)
tidy(model)

```
Insignificant variables: arm, term30, amount

Arm: interesting that arm is not significant since my intuition is that people
who obtain an arm loan with a lower initial payment would cause higher defaults
if the arm causes mortgage payments to increase from its initial value

term30: may be insignificant for a few reasons. First, the amount of records
with 15 years is low compared the entire data set. Second, borrowers with 15 year
loans may be more financially fit than 30 year borrowers. Third, the borrow can
refinance to a 30 years to lower payments if needed.

amount: I am surprised amount is insignificant as my intuition would be borrowers
with high loan amounts would have a higher likelihood to default











































