---
title: "Question 2"
author: "Robert Priolo"
date: "4/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#initial setup
rm(list=ls())
t.critical <- function(alpha, df, tail = "two") {
  if(tail == "two"){
    result <- qt(1-alpha/2  , df)
  } else if(tail == "right"){
    result <- qt(1-alpha  , df)
  } else if(tail=="left"){
    result <- qt(alpha  , df)
  }
  return(result)
}

#data
library(POE5Rdata)
data("vegas5")
attach(vegas5)

#libraries
library(broom)

```



#Question 2

Estimate a multiple linear regression model that includes all the main effects only (i.e., no
interactions nor higher order terms). We will use this model as a baseline. Comment on
the statistical and economic significance of your estimates. Also, make sure to provide an
interpretation of your estimates.

```{r}

model<- lm(default~arm+refinance+lien2+term30+underwater+ltv+rate+amount+fico)
summary(model)
tidy(model)

```
Insignificant variables: arm, term30, amount

Arm: interesting that arm is not significant since my intuition is that people
who obtain an arm loan with a lower initial payment would cause higher defaults
if the arm causes mortgage payments to increase from its initial value

term30: may be insignificant for a few reasons. First, the amount of records
with 15 years is low compared the entire data set. Second, borrowers with 15 year
loans may be more financially fit than 30 year borrowers. Third, the borrow can
refinance to a 30 years to lower payments if needed.

amount: I am surprised amount is insignificant as my intuition would be borrowers
with high loan amounts would have a higher likelihood to default